### 2022-12-05 lmcs/priority-gv.tex
Further changes

Types $\ty{\typrod{T}{U}}$, $\ty{\tyunit}$, $\ty{\tysum{T}{U}}$, and $\ty{\tyvoid}$ are the standard linear $\lambda$-calculus product type, unit type, sum type, and empty type.
Type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$ is the standard linear function type, annotated with \emph{priority bounds} $\cs{p},\cs{q}\in\mathbb{N}\cup\{\cs{\pbot},\cs{\ptop}\}$.
Every session type is also a type.
Given a function with type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$, $\cs{p}$ is a \emph{lower bound} on the priorities of the endpoints captured by the body of the function, and $\cs{q}$ is an \emph{upper bound} on the priority of the communications that take place as a result of applying the function. The type of \emph{pure functions} $\ty{\tylolli{T}{U}}$, \ie those which perform no communications, is syntactic sugar for $\ty{\tylolli[\cs{\ptop},\cs{\pbot}]{T}{U}}$.

@@

\subsubsection*{Terms Typing}
%\Cref{fig:pgv-typing} gives the typing rules for PGV.
Typing rules for terms are at the top of \cref{fig:pgv-typing}.
Terms are typed by a judgement $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ stating that ``a term $\tm{M}$ has type $\ty{T}$ and an upper bound on its priority $\cs{p}$ under the typing environment $\ty{\Gamma}$''. Typing for the linear $\lambda$-calculus is standard. Linearity is ensured by splitting environments on branching rules, requiring that the environment in the variable rule consists of just the variable, and the environment in the constant and unit rules are empty. Constants $\tm{K}$ are typed using type schemas, and embedded using \LabTirName{T-Const} (mid of \cref{fig:pgv-typing}). The typing rules treat \emph{all variables} as linear resources, even those of non-linear types such as $\ty{\tyunit}$. However, the rules can easily be extended to allow values with unrestricted usage~\cite{wadler14}.

@@

In PGV, we keep track of a lower and upper bound on the priorities of a term, \ie, while evaluating the term, when it starts communicating, and it finishes, respectively. The upper bound is written on the sequent, whereas the lower bound is approximated from the typing environment. Typing rules for sequential constructs enforce sequentiality, \eg, the typing for $\tm{\andthen{M}{N}}$ has a side condition which requires that the upper bound of $\tm{M}$ is smaller than the lower bound of $\tm{N}$, \ie, $\tm{M}$ finishes before $\tm{N}$ starts. The typing rule for $\tm{\new}$ ensures that both endpoints of a channel share the same priorities. Together, these two constraints guarantee deadlock freedom.

To illustrate this, let's go back to the deadlocked program in \cref{ex:deadlock}. Crucially, it composes the terms below in parallel. While each of these terms itself is well typed, they impose opposite conditions on the priorities, so their composition is ill-typed. (We omit the priorities on $\ty{\tyends}$ and $\ty{\tyendr}$.)
\begin{mathpar}
  \inferrule*{
    \tseq[\cs{o'}]
    {\tmty{y'}{\tyrecv[\cs{o'}]{\tyunit}{\tyendr}}}
    {\recv\;{y'}}
    {\typrod{\tyunit}{\tyendr}}
    \\
    \tseq[\cs{p}]
    {\tmty{x}{\tysend[\cs{o}]{\tyunit}{\tyends}},\tmty{y'}{\tyendr}}
    {\letbind{x}{\send\;{\pair{\unit}{x}}}{\dots}}
    {\tyunit}
    \and
    \cs{o'}<\cs{o}
  }{%
    \tseq[\cs{p}]
    {%
      \tmty{x}{\tysend[\cs{o}]{\tyunit}{\tyends}},
      \tmty{y'}{\tyrecv[\cs{o'}]{\tyunit}{\tyendr}}
    }
    {\letpair{\unit}{y'}{\recv\;{y'}}{\letbind{x}{\send\;{\pair{\unit}{x}}}{\dots}}}
    {\tyunit}
  }

@@

@@

Let us consider an example, adapted from Lindley and Morris~\cite{lindleymorris15}, in which a term returns an endpoint linked to an echo server. The echo server receives a value and sends it back unchanged. Consider the program which creates a new channel, with endpoints $\tm{x}$ and $\tm{x'}$, spawns off an echo server listening on $\tm{x}$, and then returns $\tm{x'}$:
\[
  \begin{array}{ll}
    \tm{\main}
    & \tm{\letpair{x}{x'}{\new}{}}
    \\
    & \tm{\andthen{\spawn\;(\lambda\unit.\echo_{x})}{x'}}
  \end{array}
  \hfill\qquad\hfill
  \begin{array}{lcl}
    \tm{\echo_x}
    & \defeq
    & \tm{\letpair{y}{x}{\recv\;{x}}{}}
    \\
    &
    & \tm{\letbind{x}{\send\;\pair{y}{x}}{\close\;x}}
  \end{array}
\]

>>>>

Types $\ty{\typrod{T}{U}}$, $\ty{\tyunit}$, $\ty{\tysum{T}{U}}$, and $\ty{\tyvoid}$ are the standard linear $\lambda$-calculus product type, unit type, sum type, and empty type.
Type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$ is the standard linear function type, annotated with \emph{priority bounds} $\cs{p},\cs{q}\in\mathbb{N}\cup\{\cs{\pbot},\cs{\ptop}\}$.
Every session type is also a type.
Given a function with type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$, $\cs{p}$ is a \emph{lower bound} on the priorities of the endpoints captured by the body of the function, and $\cs{q}$ is an \emph{upper bound} on the priority of the communications that take place as a result of applying the function. The type of \emph{pure functions} $\ty{\tylolli{T}{U}}$, \ie those which perform no communications, is syntactic sugar for $\ty{\tylolli[\cs{\ptop},\cs{\pbot}]{T}{U}}$. The lower bound for a pure function is $\cs{\ptop}$ as pure functions never start communicating. For similar reasons, the upper bound for a pure function is $\cs{\pbot}$.

@@

\subsubsection*{Terms Typing}
%\Cref{fig:pgv-typing} gives the typing rules for PGV.
Typing rules for terms are at the top of \cref{fig:pgv-typing}.
Terms are typed by a judgement $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ stating that ``a term $\tm{M}$ has type $\ty{T}$ and an upper bound on its priority $\cs{p}$ under the typing environment $\ty{\Gamma}$''. Typing for the linear $\lambda$-calculus is standard. Linearity is ensured by splitting environments on branching rules, requiring that the environment in the variable rule consists of just the variable, and the environment in the constant and unit rules are empty. Constants $\tm{K}$ are typed using type schemas, and embedded using \LabTirName{T-Const} (mid of \cref{fig:pgv-typing}). The typing rules treat \emph{all variables} as linear resources, even those of non-linear types such as $\ty{\tyunit}$, though they can easily be extended to allow values with unrestricted usage~\cite{wadler14}.

@@

In PGV, we keep track of a lower and upper bound on the priorities of a term, \ie, while evaluating the term, when it starts communicating, and when it finishes, respectively. The upper bound is written on the sequent and the lower bound is approximated from the typing environment, \eg, for $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ the upper bound is $\cs{p}$ and the lower bound is at least $\minpr(\ty{\Gamma})$. The latter is \emph{correct} because a term cannot communicate at a priority earlier than the earliest priority amongst the channels it has access to. It is an \emph{approximation} on function terms, as these can ``skip'' communication by returning the corresponding channel unused. However, linearity prevents such functions from being used in well-typed configurations: once the unused channel's priority has passed, it can no longer be used.

Typing rules for sequential constructs enforce sequentiality, \eg, the typing for $\tm{\andthen{M}{N}}$ has a side condition which requires that the upper bound of $\tm{M}$ is smaller than the lower bound of $\tm{N}$, \ie, $\tm{M}$ finishes before $\tm{N}$ starts. The typing rule for $\tm{\new}$ ensures that both endpoints of a channel share the same priorities. Together, these two constraints guarantee deadlock freedom.

To illustrate this, let's go back to the deadlocked program in \cref{ex:deadlock}. Crucially, it composes the terms below in parallel. While each of these terms itself is well typed, they impose opposite conditions on the priorities, so connecting $\ty{x}$ to $\ty{x'}$ and $\ty{y}$ to $\ty{y'}$ using \LabTirName{T-Res} is ill-typed, as there is no assignment to $\cs{o}$ and $\cs{o'}$ that can satsify both $\cs{o}<\cs{o'}$ and $\cs{o'}<\cs{o}$. (We omit the priorities on $\ty{\tyends}$ and $\ty{\tyendr}$.)
\begin{mathpar}
  \inferrule*{
    \tseq[\cs{o'}]
    {\tmty{y'}{\tyrecv[\cs{o'}]{\tyunit}{\tyendr}}}
    {\recv\;{y'}}
    {\typrod{\tyunit}{\tyendr}}
    \\
    \tseq[\cs{p}]
    {\tmty{x}{\tysend[\cs{o}]{\tyunit}{\tyends}},\tmty{y'}{\tyendr}}
    {\letbind{x}{\send\;{\pair{\unit}{x}}}{\dots}}
    {\tyunit}
    \and
    \cs{o'}<\cs{o}
  }{%
    \tseq[\cs{p}]
    {%
      \tmty{x}{\tysend[\cs{o}]{\tyunit}{\tyends}},
      \tmty{y'}{\tyrecv[\cs{o'}]{\tyunit}{\tyendr}}
    }
    {\letpair{\unit}{y'}{\recv\;{y'}}{\letbind{x}{\send\;{\pair{\unit}{x}}}{\dots}}}
    {\tyunit}
  }

@@

For simplicity, we assume priority annotations are not inferred, but provided as an input to type checking. However, for any term, priorities can be inferred, \eg, by using the topological ordering of the directed graph where the vertices are the priority meta-variables and the edges are the inequality constraints between the priority meta-variables in the typing derivation.

@@

Let us consider an example, adapted from Lindley and Morris~\cite{lindleymorris15}, in which a term returns an endpoint linked to an echo server. The echo server receives a value and sends it back unchanged. Consider the program which creates a new channel, with endpoints $\tm{x}$ and $\tm{x'}$, spawns off an echo server listening on $\tm{x}$, and then returns $\tm{x'}$:
\[
  \begin{array}{ll}
    \tm{\main}
    & \tm{\letpair{x}{x'}{\new\;\unit}{}}
    \\
    & \tm{\andthen{\spawn\;(\lambda\unit.\echo_{x})}{x'}}
  \end{array}
  \hfill\qquad\hfill
  \begin{array}{lcl}
    \tm{\echo_x}
    & \defeq
    & \tm{\letpair{y}{x}{\recv\;{x}}{}}
    \\
    &
    & \tm{\letbind{x}{\send\;\pair{y}{x}}{\close\;x}}
  \end{array}
\]
